---
title: 'Assignment: Query XML'
author: Agnibha Chatterjee
date: Fall 2023
---

## Question 3
Add a DTD to the XML that validates the XML
```{q3, }
<!DOCTYPE Root [
        <!ELEMENT Root (Customers, Orders)*>
        <!ELEMENT Customers (Customer)*>
        <!ELEMENT Customer (CompanyName, ContactName, ContactTitle, Phone, FullAddress, Fax)*>
        <!ATTLIST Customer
                CustomerID ID #REQUIRED>
        <!ELEMENT CompanyName (#PCDATA)>
        <!ELEMENT ContactName (#PCDATA)>
        <!ELEMENT ContactTitle (#PCDATA)>
        <!ELEMENT Phone (#PCDATA)>
        <!ELEMENT FullAddress (Address, City, Region, PostalCode, Country)*>
        <!ELEMENT Address (#PCDATA)>
        <!ELEMENT City (#PCDATA)>
        <!ELEMENT Region (#PCDATA)>
        <!ELEMENT PostalCode (#PCDATA)>
        <!ELEMENT Country (#PCDATA)>
        <!ELEMENT Fax (#PCDATA)>
        <!ELEMENT Orders (Order)*>
        <!ELEMENT Order (CustomerID, EmployeeID, OrderDate, RequiredDate, ShipInfo)*>
        <!ELEMENT CustomerID (#PCDATA)>
        <!ELEMENT EmployeeID (#PCDATA)>
        <!ELEMENT OrderDate (#PCDATA)>
        <!ELEMENT RequiredDate (#PCDATA)>
        <!ELEMENT ShipInfo (ShipVia, Freight, ShipName, ShipAddress, ShipCity, ShipRegion, ShipPostalCode, ShipCountry)*>
        <!ATTLIST ShipInfo
                ShippedDate CDATA #IMPLIED>
        <!ELEMENT ShipVia (#PCDATA)>
        <!ELEMENT Freight (#PCDATA)>
        <!ELEMENT ShipName (#PCDATA)>
        <!ELEMENT ShipAddress (#PCDATA)>
        <!ELEMENT ShipCity (#PCDATA)>
        <!ELEMENT ShipRegion (#PCDATA)>
        <!ELEMENT ShipPostalCode (#PCDATA)>
        <!ELEMENT ShipCountry (#PCDATA)>
        ]>
```


## Question 4
Load the XML with the DTD into R using validation.
```{r q4, }
library(XML)
currDir <- getwd()
xmlFile <- "CustomersAndOrders.xml"
pathToFile <- paste0(currDir, .Platform$file.sep, xmlFile)

# Parsing the XML file
# trim is used to trim whitespaces
parsedXML <- xmlParse(pathToFile, validate = T, trim = T)
```

## Question 5
Execute an XPath expression that returns the names of all customers that do not live in "MA"?
```{r q5,}
xpathEx <- "/Root/Customers/Customer[FullAddress/Region != 'MA']/ContactName"
customerWhoDontLiveInMA <- xpathSApply(parsedXML, xpathEx, xmlValue)
print(customerWhoDontLiveInMA)
```
## Question 6
Using the result returned in (5), count how many customers there are.
```{r q6,}
print(paste('Total no.of customers who do not live in MA = ', length(customerWhoDontLiveInMA)))
```

## Question 7
Using a combination of R and XPath, calculate the total amount paid for freight for all orders placed by customer with ID "GREAL".
```{r q7,}
xpathEx <- "/Root/Orders/Order[CustomerID='GREAL']/ShipInfo/Freight"
  
customerWithIdGreal <- xpathSApply(parsedXML, xpathEx, xmlValue)
customerWithIdGreal <- as.double(customerWithIdGreal)

print(paste('Total Freight = ', sum(customerWithIdGreal)))
```


## Question 8
Using a combination of R and XPath, calculate the average amount paid for freight for all orders shipped to the USA.
```{r q8,}
xpathEx <- "/Root/Orders/Order[ShipInfo/ShipCountry = 'USA']/ShipInfo/Freight"
  
ordersShippedToUS <- xpathSApply(parsedXML, xpathEx, xmlValue)
ordersShippedToUS <- as.double(ordersShippedToUS)

print(paste('Average Freight = ', mean(ordersShippedToUS)))
```